---
title: "Combine County Summaries Across Voter Files"
author: "Earl F Glynn<br><small>watchdoglab.substack.com/</small>"
date: "<small>`r Sys.Date()`</small>"
output:
  html_document:
    code_download: true
    theme: cerulean
    toc: yes
    toc_depth:  3
    toc_float:
      collapsed:  yes
      smooth_scroll: yes
    number_sections: yes
    code_folding:  show
---

```{r setup, echo = FALSE}
# http://biostat.mc.vanderbilt.edu/wiki/Main/KnitrHtmlTemplate
require(Hmisc)    # provides knitrSet and other functions
knitrSet(lang = 'markdown',   # If using blogdown: knitrSet(lang='blogdown')
         fig.align = 'left',
         w = 6.5,
         h = 4.5,
         cache = FALSE)
```

`r hidingTOC(buttonLabel = "Outline")`

```{r startYourEngines, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  comment = NA)

time.1 <- Sys.time()
```

# Setup {.tabset .tabset-fade .tabset-pills}

## {.active}

## Constants

```{r}
EXCEL_LIMIT <- 2^20
```

## Packages

```{r Packages}
library(tidyverse)
library(lubridate)
```

Display formatting

```{r Display}
library(kableExtra)  # kable_styling
```

I/O

```{r IO}
library(readxl)      # read_xlsx
library(readr)       # write_csv
library(writexl)     # write_xlsx
```

## Helper functions

```{r Helpers}
Show <- function(data, caption="", bigMark="",
                 height = NULL, width = NULL, ...)
{
  data                                       |>
  kable("html", caption=caption,
        format.args=list(big.mark=bigMark))  |>
  kable_styling(bootstrap_options=c("striped", "bordered", "condensed"),
                position="left",
                full_width=FALSE, ...)       |>
  scroll_box(height = height, width = width)
}
```

# List of summary files

Voter files considered here are over the same time span as the online SoS dashboard files.

Use these voter file summaries as supplementary data since many months of data are missing online.

Sadly, deal with differences by file using brute force here.

```{r Names}
summaryFile <-
  read_lines("File-List-First-Look-Voter-Files.txt") |>
  word(1)

summaryFile
```

```{r}
length(summaryFile)
```

```{r}
fileDate <- str_sub(summaryFile, 35, 44)
fileDate
```

# Helper function

A variety of formats

```{r readFile}
readFile <- function(index)
{
   Date <- ymd(fileDate[index])
   cat(index, Date, "\n")

  raw <-
    read_xlsx(summaryFile[index]) |>
    head(105)                       # Remove state summary in files 2023-02-13 and 2024-01-03

  if ( ( Date >= ymd("2021-09-02") ) && ( Date <= ymd("2023-08-25") ) )
  {
    d <- raw |> select(County,
                       Voters = Total, Active = A, Inactive = I,
                       Democratic, Libertarian, Republican, Unaffiliated)
  }

  if (Date == ymd("2024-01-03"))
  {
    d <- raw |> select(County,
                       Voters, Active, Inactive,
                       Democratic, Libertarian, Republican, Unaffiliated = Unaffilated)
  }

  if  ( Date == ymd("2024-05-02") )
  {
    d <- raw |> select(County,
                       Voters, Active, Inactive,
                       Democratic, Libertarian, Republican, NoLabels, Unaffiliated = Unaffilated)
  }

  if  ( Date >= ymd("2024-06-24") )
  {
    d <- raw |> select(County,
                       Voters, Active, Inactive,
                       Democratic, Libertarian, Republican, NoLabels, UnitedKansas,
                       Unaffiliated = Unaffilated)
  }

  # Create state totals as "KANSAS" County
  d <-
    bind_rows(
               d,
               tibble(
                       County       = "KANSAS",
                       Voters       = sum(d$Voters),
                       Active       = sum(d$Active),
                       Inactive     = sum(d$Inactive),
                       Democratic   = sum(d$Democratic,   na.rm = TRUE),
                       Republican   = sum(d$Republican,   na.rm = TRUE),
                       Libertarian  = sum(d$Libertarian,  na.rm = TRUE),
                       NoLabels     = sum(d$NoLabels,     na.rm = TRUE),
                       UnitedKansas = sum(d$UnitedKansas, na.rm = TRUE),
                       Unaffiliated = sum(d$Unaffiliated, na.rm = TRUE),
                     )
              )

  # Compute additional values
  d <-
    d                                                       |>
    mutate(
            Date   = Date,
     #       Active = Voters - Inactive,
            InactivePercent = round(100* Inactive / Voters, 2)
           )                                                |>
    relocate(Date,            .after = "County")            |>
    relocate(Active,          .after = "Voters")            |>
    relocate(InactivePercent, .after = "Inactive")

  return(d)
}
```

# Read all files

```{r}
countySummary <-
  1:length(fileDate) |>
  purrr::map_dfr(readFile)

glimpse(countySummary)
```

```{r}
tail(countySummary, 3) |> Show()
```

```{r}
countySummary |> filter(County == "Sumner")  |> Show(bigMark = ",")
```

```{r}
countySummary |> filter(County == "Kansas")  |> Show(bigMark = ",")
```

```{r}
countySummary <-
  countySummary |>
  rename(Name = County)  |>
  mutate(
          Geography = if_else(Name == "KANSAS", "State", "County")
        )                |>
  relocate(Geography, Name) |>
  arrange(Geography, Name, Date)
```

```{r}
write_xlsx(countySummary,
           paste0("Kansas-Voter-Counts-",
                  head(fileDate,1),
                  "-thru-",
                  tail(fileDate,1),
                  ".xlsx"))
```

# Epilog {.tabset .tabset-fade .tabset-pills}

## {.active}

## Session Info

```{r devtoolsSessionInfo}
devtools::session_info()
```

</div>

```{r epilogDeltaTime, echo=FALSE}
time.2 <- Sys.time()
processingTime <- paste("Processing time:", sprintf("%.1f",
                        as.numeric(difftime(time.2,
                                            time.1, units="secs"))), "secs\n")
```

`r processingTime`
`r format(time.2, "%Y-%m-%d %H%M")`

