---
title: "Kansas Inactive Voter Percentages from Voter Files -- All Years"
author: "Earl F Glynn<br><small>watchdoglab.substack.com/</small>"
date: "<small>`r Sys.Date()`</small>"
output:
  html_document:
    code_download: true
    theme: cerulean
    toc: yes
    toc_depth:  3
    toc_float:
      collapsed:  yes
      smooth_scroll: yes
    number_sections: yes
    code_folding:  show
---

```{r setup, echo = FALSE}
# http://biostat.mc.vanderbilt.edu/wiki/Main/KnitrHtmlTemplate
require(Hmisc)    # provides knitrSet and other functions
knitrSet(lang = 'markdown',   # If using blogdown: knitrSet(lang='blogdown')
         fig.align = 'left',
         w = 6.5,
         h = 4.5,
         cache = FALSE)
```

`r hidingTOC(buttonLabel = "Outline")`

```{r startYourEngines, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  comment = NA)

time.1 <- Sys.time()
```

# Setup {.tabset .tabset-fade .tabset-pills}

## {.active}

## Constants

```{r Constants}
FILE_PREFIX   <- "x"    # no spaces
```

```{r}
EXCEL_LIMIT <- 2^20
```

## Packages

```{r Packages}
library(tidyverse)
library(lubridate)
```

Display formatting

```{r Display}
library(kableExtra)  # kable_styling
#library(IRdisplay)   # display_html (needed in Jupyter notebooks)
```

I/O

```{r IO}
library(readxl)      # read_xlsx
library(readr)       # write_csv
library(writexl)     # write_xlsx
```

 Graphics

```{r Graphics}
library(RColorBrewer) # brewer.pal
library(scales)       # comma, comma_format
library(gplots)       # heatmap.2
```

Utilities

```{r Utilities}
library(downloader)  # download
library(tools)       # md5sum
```

## ggplot theme

[ggplot Themes Gallery](https://www.datanovia.com/en/blog/ggplot-themes-gallery/)

```{r ggplotTheme}
theme_set(theme_minimal() +

          theme(axis.text             = element_text(size = 12),
                axis.title            = element_text(size = 14),

                legend.position       = "bottom",

                plot.caption          = element_text(hjust = c(0.0,1.0),
                                                     size = 10),
                plot.caption.position = "plot",

                plot.title.position   = "plot",

                strip.background      = element_rect(fill = "aliceblue"),
                strip.text            = element_text(size = 14),

                title                 = element_text(size = 14)))

COLOR_BAR     <- "skyblue"
COLOR_OUTLINE <- "grey80"
```

## Helper functions

```{r Helpers}
Show <- function(data, caption="", bigMark="",
                 height = NULL, width = NULL, ...)
{
  data                                       |>
  kable("html", caption=caption,
        format.args=list(big.mark=bigMark))  |>
  kable_styling(bootstrap_options=c("striped", "bordered", "condensed"),
                position="left",
                full_width=FALSE, ...)       |>
  scroll_box(height = height, width = width)
}
```

# Kansas inactive rates based on voter files

Historical data that precedes online SoS Dashboard Summary Files

File from `C:\Data\Kansas\RawData\Voter-Election-Codes-2-Lookup-Voter-Name-Raw-Data\Names\Kansas`

```{r}
voterFiles <- 
  read_xlsx("Kansas-Voters-Source-File-Summary-Edited.xlsx") |>
  select(Source, InactivePercent = `%Inactive`)              |>
  mutate(
          Date = ymd(Source),
          InactivePercent = 100 * InactivePercent
        )

glimpse(voterFiles)
```

```{r}
c(min(voterFiles$Date), max(voterFiles$Date))
```

```{r}
cycles <- seq( from = ymd("2005-01-01"), to = ymd("2027-01-01"), by = "2 years")
cycles
```

```{r}
cycleLabels <- seq( from = ymd("2008-01-01"), to = ymd("2026-01-01"), by = "2 years")
cycleLabels
```

```{r}
plotCaptionLeft  <- "Source: Kansas Secretary of State Voter Files [2006 - 2026]"
plotCaptionRight <- paste("efg ", format(Sys.time(), "%Y-%m-%d"))
stateColor   <- "royalblue"  
```     

```{r Kansas-Inactive-Voters-Historical, fig.width = 10, fig.height = 6}    
 ggplot() +
     geom_point(data = voterFiles,
               aes(x = Date,
                   y = InactivePercent),
               color = stateColor, 
               fill= stateColor,
               size = 3,
               shape = 25)                              +
   scale_y_continuous(
                       limits = c(4, 15),
                       breaks = c( 0,  5, 10, 15, 20, 
                                  25, 30, 35, 40),
                       labels = c( "0%",  "5%", "10%", "15%", "20%", 
                                   "25%", "30%", "35%", "40%")  
                     )                                  +
   
  geom_vline(xintercept = cycles,
              color     = "black",
              linewidth = 1,
              linetype  = "solid"
             )                                          +
  
  annotate("text", x = cycleLabels,
                   y = Inf, 
                   label = paste0(str_sub(cycleLabels, 1, 4), "\nCycle"),
            color = "black", hjust = 0.5, 
            vjust = 1, size = 4)                        +
  
  scale_x_date(limits = c(ymd("2006-05-01"), ymd("2026-06-01")),
                 date_breaks = "1 year",
                 date_labels = "%Y",
                 expand = c(0,0))                       +
  labs(title    = "Percent Inactive Voters in Kansas over Twenty Years (May 2006 - May 2026)",
       subtitle = "Based on Kansas voter files",
       x = "Date",
       y = "Inactive Voters [%]",
       caption = c(plotCaptionLeft, plotCaptionRight))  +
 theme(plot.title         = element_text(size = 14, color = stateColor, face = "bold"),
         plot.subtitle      = element_text(size = 14, color = "grey"),
         axis.ticks.x       = element_line(),
         panel.grid.minor.x = element_blank(),
         panel.grid.major.x = element_line(linewidth = 0.5, color = "grey"),
        # panel.grid.minor.y = element_blank(),
         panel.grid.major.y = element_line(linewidth = 0.5, color = "grey"))        
```

# Epilog {.tabset .tabset-fade .tabset-pills}

## {.active}

## Session Info

```{r devtoolsSessionInfo}
devtools::session_info()
```

</div>

```{r epilogDeltaTime, echo=FALSE}
time.2 <- Sys.time()
processingTime <- paste("Processing time:", sprintf("%.1f",
                        as.numeric(difftime(time.2,
                                            time.1, units="secs"))), "secs\n")
```

`r processingTime`
`r format(time.2, "%Y-%m-%d %H%M")`

