---
title: "Load and Plot Kansas Inactive Voter Percentages by County"
author: "Earl F Glynn<br><small>watchdoglab.substack.com/</small>"
date: "<small>`r Sys.Date()`</small>"
output:
  html_document:
    code_download: true
    theme: cerulean
    toc: yes
    toc_depth:  3
    toc_float:
      collapsed:  yes
      smooth_scroll: yes
    number_sections: yes
    code_folding:  show
---

```{r setup, echo = FALSE}
# http://biostat.mc.vanderbilt.edu/wiki/Main/KnitrHtmlTemplate
require(Hmisc)    # provides knitrSet and other functions
knitrSet(lang = 'markdown',   # If using blogdown: knitrSet(lang='blogdown')
         fig.align = 'left',
         w = 6.5,
         h = 4.5,
         cache = FALSE)
```

`r hidingTOC(buttonLabel = "Outline")`

```{r startYourEngines, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  comment = NA)

time.1 <- Sys.time()
```

# Setup {.tabset .tabset-fade .tabset-pills}

## {.active}

## Constants

```{r Constants}
FILE_PREFIX   <- "x"    # no spaces
```

```{r}
EXCEL_LIMIT <- 2^20
```

## Packages

```{r Packages}
library(tidyverse)
library(lubridate)
```

Display formatting

```{r Display}
library(kableExtra)  # kable_styling
#library(IRdisplay)   # display_html (needed in Jupyter notebooks)
```

I/O

```{r IO}
library(readxl)      # read_xlsx
library(readr)       # write_csv
library(writexl)     # write_xlsx
```

 Graphics

```{r Graphics}
library(RColorBrewer) # brewer.pal
library(scales)       # comma, comma_format
library(gplots)       # heatmap.2
```

Utilities

```{r Utilities}
library(downloader)  # download
library(tools)       # md5sum
```

## ggplot theme

[ggplot Themes Gallery](https://www.datanovia.com/en/blog/ggplot-themes-gallery/)

```{r ggplotTheme}
theme_set(theme_minimal() +

          theme(axis.text             = element_text(size = 12),
                axis.title            = element_text(size = 14),

                legend.position       = "bottom",

                plot.caption          = element_text(hjust = c(0.0,1.0),
                                                     size = 10),
                plot.caption.position = "plot",

                plot.title.position   = "plot",

                strip.background      = element_rect(fill = "aliceblue"),
                strip.text            = element_text(size = 14),

                title                 = element_text(size = 14)))

COLOR_BAR     <- "skyblue"
COLOR_OUTLINE <- "grey80"
```

## Helper functions

```{r Helpers}
Show <- function(data, caption="", bigMark="",
                 height = NULL, width = NULL, ...)
{
  data                                       |>
  kable("html", caption=caption,
        format.args=list(big.mark=bigMark))  |>
  kable_styling(bootstrap_options=c("striped", "bordered", "condensed"),
                position="left",
                full_width=FALSE, ...)       |>
  scroll_box(height = height, width = width)
}
```

# Voter file summaries

```{r}
voterFileSummary <- 
  read_xlsx("../Stats-from-Voter-Files/Kansas-Voter-Counts-2021-09-02-thru-2026-05-04.xlsx") |>
  select(Geography, Name, Date, InactivePercent) |>
  mutate(Date = ymd(Date))

glimpse(voterFileSummary)
```

```{r}
c(min(voterFileSummary$Date), max(voterFileSummary$Date))
```

# SoS Summary File Background

Assumes completion of `1-Download-Monthly-County-Inactive-Lists.html` to download files.

From KS SoS Voter Registration Statistics Dashboard
https://www.sos.ks.gov/elections/voter-registration-statistics.html

Inspection of files across months and years shows inconsistencies among files.

# List of SoS files

```{r}
file <- 
  tibble(
          Filename = 
            list.files( 
                        path = "Data",
                        pattern = glob2rx("*-Percentage-of-Inactive-Voters-by-County.xlsx")
                      )
        ) |>
  mutate(
          Date   = ymd(str_sub(Filename, 1, 10)),
          Rows   = NA_integer_,
          Cols   = NA_integer_,
          header = NA_character_,
          types  = NA_character_
        )
 
dim(file)
```

# First Look

How many rows and columns

```{r}
for (i in 1:nrow(file))
{
  cat(i, as.character(file$Date[i]), "\n")
  set <- read_xlsx(paste0("Data/", file$Filename[i])) 
  file$Rows[i]   <- nrow(set)
  file$Cols[i]   <- ncol(set)
  file$header[i] <- str_flatten(names(set), collapse = "|")
  file$types     <- str_flatten(sapply(set, typeof), collapse = "|")
}
```
Introuction of new name "...1" caused by missing County name in "Old" format files

```{r}
file |> Show()
```

# Deal with variety of formats

Arbitrary changes made over time to the same data.

Introduce consistency among the files. Standardize on names used starting in March 2026.

The number of Rows and Cols identifies file format.

```{r}
readSoSFile <- function(index, file, DataPath = "Data/")
{
  Filename <- paste0(DataPath, file$Filename[index])
  DataDate <- file$Date[index]
  
  cat(index, as.character(DataDate), "\n")
 
  set <- read_xlsx(Filename)    
  
  # "Old" format:  2021-06-01 through 2025-12-01
  if ( (nrow(set) == 108) && (ncol(set) == 4) )
  {
    # Assign "standard" names
    names(set) <- c("County", "Total", "Inactive", "InactivePercent")
    
    set$County[nrow(set)] <- "KANSAS"    # label for stat row is missing in original
  }
  
  # "New" format 1:  2026-01-01 only known case
  if ( (nrow(set) == 107) && (ncol(set) == 5) )
  {
    # Assign "standard" names
    names(set) <- c("Date", "County", "Total", "Inactive", "InactivePercent")
    
    set <- set |> select(-Date)  # Remove since file datestamp added below for all
    
    set$County[nrow(set)] <- "KANSAS"    # label for stat row is missing in original
  }
  
  # "New" format 2:  2026-02-01 only known case
  if ( (nrow(set) == 106) && (ncol(set) == 6) )
  {
    # Assign "standard" names
    names(set) <- c("Date", "County", "Active", "Inactive", "Total", "InactivePercent")
    set <- set |> select(-Date)  # Remove since file datestamp added below for all
    set$County[nrow(set)] <- "KANSAS"    # Replace "Grand Total"
  }
  
  # "New" format 3:  2026-03-01 and later
  if ( (nrow(set) == 106) && (ncol(set) == 5) )
  {
    # Assign "standard" names
    names(set) <- c("County", "Active", "Inactive", "Total", "InactivePercent")
    set$County[nrow(set)] <- "KANSAS"    # Replace "Grand Total"
    
    # This is needed for 2026-05-01 data; unclear why character numbers in that file
    set <- 
      set |>
      mutate(
              Active   = as.numeric(Active),
              Inactive = as.numeric(Inactive)
            )
    
  }
  
  # Add DataDate to this batch
  set <-
    set                     |>
    filter(!is.na(County))  |>  # get rid of blank lines
    mutate(Date = DataDate) |>
    relocate(Date)
  
  set$County[nrow(set)] <- "KANSAS"    # label for stat row is missing in "old";
                                       # "Grand Total" in "new"
 
  invisible(set) 
}
```

# Explore 

```{r}
index <- nrow(file)
index <- 29
Filename <- paste0("Data/", file$Filename[index])
DataDate <- file$Date[index]
c(Filename, as.character(DataDate))
```

```{r}
inactive <- readSoSFile(index, file)
```

```{r}
glimpse(inactive)
```

```{r}
inactive |> head() |> Show()
```

```{r}
inactive |> tail() |> Show()
```

# Long format

## Combine all files

Long Format shows all data in original files 

```{r}
KansasInactive <- 
  1:nrow(file)                      |>
  purrr::map_dfr(readSoSFile, file) |>
  arrange(County, Date)             |>
  mutate(InactivePercent = 100 * InactivePercent)  |>
  select(Date, County, Total, Active, Inactive, InactivePercent)

dim(KansasInactive)
```

Change "County" to "Name" and add Geography column with values County or State

```{r}
KansasInactive <-
  KansasInactive         |>
  rename(Name = County)  |>
  mutate(
          Geography = if_else(Name == "KANSAS", "State", "County")
        )                |>
  relocate(Geography, Name) |>
  arrange(Geography, Name, Date)
```

```{r}
glimpse(KansasInactive)
```

```{r}
write_xlsx(KansasInactive, "Kansas-SoS-Inactive-Voter-Rate-Long-Format.xlsx")
```

```{r}
c(min(KansasInactive$Date), max(KansasInactive$Date))
```


Make sure all counties and state data present in all files

```{r}
table(KansasInactive$Name)
```

## High inactive rates 

```{r}
fivenum(KansasInactive$InactivePercent)
```

```{r}
boxplot(KansasInactive$InactivePercent)
```


```{r}
KansasInactive |>
  filter(InactivePercent > 20) |>
  Show()
```

# Wide format

Wide format will show only `InactivePercent`.  (Could make wide format for other quantities if needed.)

Wide format can be used for heatmap analysis showing which counties have similar patterns, and to create sparklines for all counties in Excel.

```{r}
KansasInactiveWide <-
  KansasInactive |>
  select(Geography, Name, Date, Date, InactivePercent) |>
  pivot_wider(names_from  = Date,
              values_from = InactivePercent)

dim(KansasInactiveWide)
```

```{r}
write_xlsx(KansasInactiveWide, "Kansas-SoS-Inactive-Voter-Rate-Wide-Format.xlsx")
```

## Heatmap

```{r}
inactiveMatrix <- 
  KansasInactiveWide        |>
  select(-Geography, -Name) |>
  as.matrix()

dim(inactiveMatrix)
```

```{r}
dimnames(inactiveMatrix) <-
  list(
         KansasInactiveWide$Name,
         names(KansasInactiveWide)[c(-1, -2)] |> str_sub(3, 7)
      )
```

```{r}
inactiveMatrix[1:5, 1:5]
```

```{r}
colorBreaks <- c(0, 2, 4, 6, 8, 10, 12, 14, 16, 20, 30, 40, 50, 64) 
colorBreaks
```


```{r}
colorPalette <-
  c(
     colorRampPalette(c("red",   "white"))(6)[1:5],    # skip last
     colorRampPalette(c("white", "skyblue"))(7)[2:5],  # skip first
     "grey80", "grey64", "grey32", "grey0"
   )                                        

colorPalette
```


```{r}
# layout matrix 1 = heatmap, 2 = row dendrogram, 3 = column dendrogram, 4 = key
# 0 = leaves space
lmat <- rbind( c(0, 3, 3),  # layout matrix
               c(2, 1, 1),
               c(0, 0, 4))

# Layout relative widths
lwid <- c(1, 4, 6)    # entry for each column in layout matrix 3-4-4
                      
# Layout relative heights
# relative height of the rows. must have length = nrow(lmat)
lhei <- c(0.15, 5, 0.3)    # entry for each row in layout matrix
```

```{r Inactive-Heatmap, fig.width = 16, fig.height = 20}
oldPar <- par(mar = c(5, 0, 2, 0))

heatmap.2(
           inactiveMatrix,
           scale      = "none",
           dendrogram = "row",
           Colv       = NA,         # suppress reordering of columns

           trace  = "none",         # turns off trace lines inside the heat map
           col    = colorPalette,
           breaks = colorBreaks,

           margins = c(4, 10),      # column and row names

           cexRow = 1.50,
           cexCol = 1.75,

           key.title = NA,           # suppress title
           key.xlab  = "Kansas Inactive Voter Percentage, June 2021 - May 2026 ", ,
           keysize   = 3,
           density.info = "none",    # turns off density plot inside color legend

           lmat = lmat, lwid = lwid, lhei = lhei
         )

title(main     = "Inactive Voter Percentage",
      cex.main = 3,
      sub      = "Voter File Date (yy-mm)",
      cex.sub  = 1.5)

par(oldPar)
```

# Separate State and County Plots

```{r}
stateStats   <- KansasInactive   |> filter(Geography == "State")  
stateVoters  <- voterFileSummary |> filter(Geography == "State")
stateColor   <- "royalblue"

c(min(stateStats$InactivePercent),  max(stateStats$InactivePercent))
c(min(stateVoters$InactivePercent), max(stateVoters$InactivePercent))
```

```{r}
stateStats |> 
  filter(
          InactivePercent == min(stateStats$InactivePercent) 
        )  |>
  Show()
```

```{r}
stateVoters |> 
  filter(
          InactivePercent == max(stateVoters$InactivePercent) 
        )  |>
  Show()
```


```{r}
plotCaptionLeft  <- "Source: Kansas Secretary of State Online Voter Registration Statistics Dashboard (circles); Voter Files (triangles)"
plotCaptionRight <- paste("efg ", format(Sys.time(), "%Y-%m-%d"))
```     

```{r}
countyStats  <- KansasInactive   |> filter(Geography == "County")  
countyVoters <- voterFileSummary |> filter(Geography == "County")
countyColor  <- "grey"
```

```{r KansasWithCountyLines, fig.width = 10, fig.height = 6}       
allCountiesPlot <-
  ggplot() +
    geom_line(data = countyStats,
              aes(x = Date,
                  y = InactivePercent,
                  group = Name),
              color = countyColor,
              linewidth = 1      )           +
  
#    geom_line(data = stateStats,
#              aes(x = Date,
#                  y = InactivePercent),
#              color = stateColor,
#              linewidth = 1)                 +
  
    geom_point(data = stateStats,
               aes(x = Date,
                   y = InactivePercent),
               color = stateColor,
               size = 3)                     +

    geom_point(data = stateVoters,
               aes(x = Date,
                   y = InactivePercent),
               color = stateColor, 
               fill= stateColor,
               size = 3,
               shape = 25)                   +
    
   scale_y_continuous(
                       limits = c(0, 25),
                       breaks = c( 0,  5, 10, 15, 20, 
                                  25, 30, 35, 40),
                       labels = c( "0%",  "5%", "10%", "15%", "20%", 
                                   "25%", "30%", "35%", "40%")  
                     )                                             +
  
   geom_vline(xintercept = c(
                              ymd("2023-01-01"),
                              ymd("2025-01-01")
                            ),
              color     = "black",
              linewidth = 1,
              linetype  = "solid"
             )                                                     +
  
   annotate("text", x = c(
                           -Inf, 
                           ymd("2024-01-01"),
                           Inf
                         ),
                    y = Inf, 
                    label = c(
                               "2021-2022 Election Cycle",
                               "2023-2024 Election Cycle",
                               "2025-2026 Election Cycle"
                             ),
            color = "black", hjust = c(0.0, 0.5, 1.0), 
            vjust = 1, size = 4)                                   +

   scale_x_date(limits = c(ymd("2021-04-20"), ymd("2026-05-10")),
                 date_breaks       = "3 months",
                 date_minor_breaks = "1 month",
                 date_labels = "%Y\n%b",
                 expand = c(0,0))                                  +
    
   labs(title    = "Percent Inactive Voters in Kansas (July 2021 - May 2026)",
        subtitle = "Grey lines show inactive voter rate profiles for each of the 105 Kansas counties",
        x = "Date",
        y = "Inactive Voters [%]",
        caption = c(plotCaptionLeft, plotCaptionRight))  +
   theme(plot.title         = element_text(size = 14, color = stateColor, face = "bold"),
         plot.subtitle      = element_text(size = 14, color = "grey"),
         axis.ticks.x       = element_line(),
         panel.grid.major.x = element_line(linewidth = 0.5, color = "grey"),
         panel.grid.minor.y = element_blank(),
         panel.grid.major.y = element_line(linewidth = 0.5, color = "grey"))        

print(allCountiesPlot)
```

# State only inactive rate based on voter files

```{r}
voterFiles <- 
  read_xlsx("Kansas-Voters-Source-File-Summary-Edited.xlsx") |>
  select(Source, InactivePercent = `%Inactive`)              |>
  mutate(
          Date = ymd(Source),
          InactivePercent = 100 * InactivePercent
        )

glimpse(voterFiles)
```

```{r}
c(min(voterFiles$Date), max(voterFiles$Date))
```

```{r}
cycles <- seq( from = ymd("2005-01-01"), to = ymd("2027-01-01"), by = "2 years")
cycles
```

```{r}
cycleLabels <- seq( from = ymd("2008-01-01"), to = ymd("2026-01-01"), by = "2 years")
cycleLabels
```


```{r}
plotCaptionLeft  <- "Source: Kansas Secretary of State Voter Files [2006 - 2026]"
```     

```{r KansasHistorical, fig.width = 10, fig.height = 6}    
 ggplot() +
     geom_point(data = voterFiles,
               aes(x = Date,
                   y = InactivePercent),
               color = stateColor, 
               fill= stateColor,
               size = 3,
               shape = 25)                              +
   scale_y_continuous(
                       limits = c(4, 15),
                       breaks = c( 0,  5, 10, 15, 20, 
                                  25, 30, 35, 40),
                       labels = c( "0%",  "5%", "10%", "15%", "20%", 
                                   "25%", "30%", "35%", "40%")  
                     )                                  +
   
  geom_vline(xintercept = cycles,
              color     = "black",
              linewidth = 1,
              linetype  = "solid"
             )                                          +
  
  annotate("text", x = cycleLabels,
                   y = Inf, 
                   label = paste0(str_sub(cycleLabels, 1, 4), "\nCycle"),
            color = "black", hjust = 0.5, 
            vjust = 1, size = 4)                        +
  
  scale_x_date(limits = c(ymd("2006-05-01"), ymd("2026-06-01")),
                 date_breaks = "1 year",
                 date_labels = "%Y",
                 expand = c(0,0))                       +
  labs(title    = "Percent Inactive Voters in Kansas over Twenty Years (May 2006 - May 2026)",
       subtitle = "Based on Kansas voter files",
       x = "Date",
       y = "Inactive Voters [%]",
       caption = c(plotCaptionLeft, plotCaptionRight))  +
 theme(plot.title         = element_text(size = 14, color = stateColor, face = "bold"),
         plot.subtitle      = element_text(size = 14, color = "grey"),
         axis.ticks.x       = element_line(),
         panel.grid.minor.x = element_blank(),
         panel.grid.major.x = element_line(linewidth = 0.5, color = "grey"),
        # panel.grid.minor.y = element_blank(),
         panel.grid.major.y = element_line(linewidth = 0.5, color = "grey"))        
```

# Epilog {.tabset .tabset-fade .tabset-pills}

## {.active}

## Session Info

```{r devtoolsSessionInfo}
devtools::session_info()
```

</div>

```{r epilogDeltaTime, echo=FALSE}
time.2 <- Sys.time()
processingTime <- paste("Processing time:", sprintf("%.1f",
                        as.numeric(difftime(time.2,
                                            time.1, units="secs"))), "secs\n")
```

`r processingTime`
`r format(time.2, "%Y-%m-%d %H%M")`

